\bf Model Visualization Techniques for a Social Network Model

### 2017-09-15 paper.Rnw
updates

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\linewidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.title.x = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.title.y = element_text(size = rel(1.5),
                                            face = 'plain', angle = 90),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'plain', angle = 0), 
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

@@

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are ``behind the scenes" of when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
}

\par The guiding principles of model visualization introduced in \citet{modelvis} are: 
\begin{enumerate}
\item Display the model in the data space.
\item Collections of models are more informative than singletons. 
\item Explore the fitting \textit{process} rather than just looking at the final result.
\end{enumerate}
Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

@@

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet + 
  theme(panel.grid = element_blank(), aspect.ratio = 1) +
  labs(title = "Adjacency Matrix Visualization", x = "to (alter)", y = "from (ego)")
@
%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. For this paper, we assume a simple rate \st{``function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}. In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.  %\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$. If $j = i$, no change in the network is made. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

@@

\hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
\st{XXX see line 1257 for plot. where should I put it? }
\hh{XXX Put a reference to the plot here.}

@@

<<logweibull, fig.align='center', out.width = '.99\\linewidth', fig.height = 2, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
library(reliaR)
x <- seq(-3,10,.01)
y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
qplot(x = x, y= y, geom = "line") + 
  annotate('text', x = 6, y = .25, 
        label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
  labs(x = "u", y = "density of g(u)")
@

@@

When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$. \st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function, or in our case the rate parameter, to fully describe the CTMC that is used to model network change.}
\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

@@

Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep(Rsiena), we have an accessible way to fit SAOMs to network observations.

@@

\subsection{Model Goodness-of-Fit}

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)

sims1 <- data.frame(gof1$`Period 1`$Simulations)
dat1 <- gof1$`Period 1`$Observations
dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
  mutate(outdegree = parse_number(outdegree)-1) -> dat1
p1 <- sims1 %>% 
  mutate(sim = row_number()) %>% 
  gather(outdegree, val, X1:X9) %>% 
  mutate(outdegree = parse_number(outdegree) - 1) %>% 
  filter(outdegree <= 6) %>%  
  ggplot(aes(x = outdegree, y = val)) + 
  geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
  geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
  geom_point(data = dat1, color = 'red') + 
  geom_line(data= dat1, color = 'red') + 
  geom_text(data = dat1, aes(label = val), hjust = -.5) + 
  labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
       title = "Goodness-of-Fit: Outdegree distribution period 1") + 
  ThemeNoNet

sims2 <- data.frame(gof2$`Period 1`$Simulations)
dat2 <- gof2$`Period 1`$Observations
dat2 %>% data.frame() %>% gather(indegree, val) %>% 
  mutate(indegree = parse_number(indegree)-1) -> dat2
p2 <- sims2 %>% 
  mutate(sim = row_number()) %>% 
  gather(indegree, val, X1:X9) %>% 
  mutate(indegree = parse_number(indegree) - 1) %>% 
  filter(indegree <= 6) %>%  
  ggplot(aes(x = indegree, y = val)) + 
  geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
  geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
  geom_point(data = dat2, color = 'red') + 
  geom_line(data= dat2, color = 'red') + 
  geom_text(data = dat2, aes(label = val), hjust = -.5) + 
  labs(x = "Indegree (p = 0.063)", y = "Statistic", 
       title = "Goodness-of-Fit: Indegree distribution period 1") + 
  ThemeNoNet

gridExtra::grid.arrange(p1, p2)
# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
# [1] 0.154
# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
# [1] 0.063

The \texttt{plot.sienaGOF()} function allows us to visualize this fit. In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model. Box plots and violin plots are drawn on top of each other in the figure in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines. If the red points lie "well within" the simulated values, the model is a good fit to the data. The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.

@@

<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = y, y=-x)) + 
  geom_tile(aes(fill = as.factor(value))) + 
  scale_fill_manual("i likes j", labels=c("no", "yes"), 
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) + 
  facet_wrap(~wave) + theme_bw() +
  labs(x = "j", y = "i") + 
  ThemeNoNet + 
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank()) 
@

>>>>

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0, 
                                            family="Times New Roman"), 
                  axis.title.x = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"),
                  axis.title.y = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 90,
                                            family="Times New Roman"),
                  axis.text.x.top = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"), 
                  axis.text.x.bottom = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"), 
                  axis.text.y.left = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"),
                  axis.text.y.right = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"),
                  strip.text.x = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0, 
                                            family="Times New Roman",
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = 12,
                                            face = 'plain', 
                                            angle = 90,
                                            family="Times New Roman",
                                            margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman"), 
                  strip.text.x = element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman", 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman",margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

@@

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
}

\par We get into the SAOM black box by using model visualization (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output. Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from the application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

@@

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet +
  theme(plot.title = element_text(size = 20)) + 
  labs(title = "Node-Link Diagram")
#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)

ggplot(data = toynet$edges) +
  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet + 
  theme(panel.grid = element_blank(), aspect.ratio = 1, 
        plot.title = element_text(size = 20), 
        axis.title.x.top = element_text(size = 18),
        axis.title.y.left = element_text(size = 18), 
        axis.text.x.top  = element_text(size = 18),
        axis.text.y.left  = element_text(size = 18)) +
  labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
@
%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}. In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.  %\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$. If $j = i$, no change in the network is made. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

@@

% \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
% \st{XXX see line 1257 for plot. where should I put it? }
% \hh{XXX Put a reference to the plot here.}

@@

<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
library(reliaR)
x <- seq(-3,10,.01)
y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
qplot(x = x, y= y, geom = "line") + 
  annotate('text', x = 6, y = .25, 
        label = "g(u) = exp(-(u + exp(-u)))",size=5, family = "Times New Roman")  + 
  theme(text = element_text(size = 20, family = "Times New Roman")) + 
  labs(x = "u", y = "density of g(u)")
@

@@

When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$. \st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.}

\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

@@

Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM. Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.

@@

\subsection{Model Goodness-of-Fit}\label{sec:modgof}

@@

% <<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
% load("data/ansnullpaper.rda")
% library(lattice)
% gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
% gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
% 
% sims1 <- data.frame(gof1$`Period 1`$Simulations)
% dat1 <- gof1$`Period 1`$Observations
% dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
%   mutate(outdegree = parse_number(outdegree)-1) -> dat1
% p1 <- sims1 %>% 
%   mutate(sim = row_number()) %>% 
%   gather(outdegree, val, X1:X9) %>% 
%   mutate(outdegree = parse_number(outdegree) - 1) %>% 
%   filter(outdegree <= 6) %>%  
%   ggplot(aes(x = outdegree, y = val)) + 
%   geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
%   geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
%   geom_point(data = dat1, color = 'red') + 
%   geom_line(data= dat1, color = 'red') + 
%   geom_text(data = dat1, aes(label = val), hjust = -.5) + 
%   labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
%        title = "Goodness-of-Fit: Outdegree distribution period 1") + 
%   ThemeNoNet
% 
% sims2 <- data.frame(gof2$`Period 1`$Simulations)
% dat2 <- gof2$`Period 1`$Observations
% dat2 %>% data.frame() %>% gather(indegree, val) %>% 
%   mutate(indegree = parse_number(indegree)-1) -> dat2
% p2 <- sims2 %>% 
%   mutate(sim = row_number()) %>% 
%   gather(indegree, val, X1:X9) %>% 
%   mutate(indegree = parse_number(indegree) - 1) %>% 
%   filter(indegree <= 6) %>%  
%   ggplot(aes(x = indegree, y = val)) + 
%   geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
%   geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
%   geom_point(data = dat2, color = 'red') + 
%   geom_line(data= dat2, color = 'red') + 
%   geom_text(data = dat2, aes(label = val), hjust = -.5) + 
%   labs(x = "Indegree (p = 0.063)", y = "Statistic", 
%        title = "Goodness-of-Fit: Indegree distribution period 1") + 
%   ThemeNoNet
% 
% 
% gridExtra::grid.arrange(p1, p2)
% # sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
% # [1] 0.154
% # > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
% # [1] 0.063
% 
% @

The \texttt{plot.sienaGOF()} function allows us to visualize this fit. %In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model. 
This function draws a box plots and a violin plot at each value of the statistic of interest observed in the simulations. In order to compare the distribution of the simulated network to the true data, red points connected by red lines representing the observed values are superimposed on the boxplots. If the red points lie "well within" the simulated values, the model is a good fit to the data. The plot is not intuitive for understanding network data. %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data. 
In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.

@@

<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = y, y=-x)) + 
  geom_tile(aes(fill = as.factor(value))) + 
  scale_fill_manual("i likes j", labels=c("no", "yes"), 
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) + 
  facet_wrap(~wave) + theme_bw() +
  labs(x = "node j (alter)", y = "node i (ego)") + 
  ThemeNoNet + 
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank()) 
@
